\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-07 chapters/wcble.tex
funding

\label{sec:wcble}
\todo{comments liesel}
\emph{Sections~\ref{sec:wcble/literature/contrib} and~\ref{sec:wcble/methods} were published
as part of~\textcite{VanDenKerchove2024}}

@@

\Ac{cble}, summarized in Algorithm~\ref{alg:cble}, works by training a
\textit{first-stage} classifier $\mathcal{C}(\theta,f)$
defined within a time period $[s_1:s_2]$ with a set of parameters $\theta$ and a
decision function $f(\mat{X}[:,s_1:s_2],\theta) \rightarrow
y$ outputting a classification score $y\in\mathbb{R}$ for a given epoch $\mat{X}$,
such that
\begin{equation}
  \theta = \mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
  \label{eq:cble-train}
\end{equation}
Then,
%instead of applying the classifier directly to an unseen, cropped
%testing epoch $\mat{X}[:,s_1:s_2]$ and obtaining a single score value $y$,
$f$ can be applied to all (possibly overlapping) slices of length $s_2-s_1$ of
an epoch $\mat{X}$, resulting in a vector of score values
$\mathbf{y}=[y_1\,\ldots\,y_R]^T \in\mathbb{R}^R$ such that
\begin{equation}
  y_s = f(\mat{X}[:,s:s+(s_2-s_1)],\theta)\quad \forall s\in 1,\ldots,R
	\label{eq:score}
\end{equation}
with $R = S-(s_2-s_1)$.
To leverage \ac{cble} for \ac{erp} classification, the score vectors $\mathbf{y}$ can be
arranged in matrices $\mat{Y}^\mathrm{train}\in\mathbb{R}^{N\times R}$ and $\mat{Y}^\mathrm{test}\in\mathbb{R}^{M\times R}$.
These can be further classified by training a \textit{second-stage} classifier on
$\mat{Y}^\mathrm{train}$ and class labels $\mathbf{l}^\mathrm{train}$.
However, the resulting score-over-time vectors per epoch still suffer from jitter.
For classification, we follow the approach of \cite{Mowla2017}, using a
maximum-level hierarchical Daubechies-4 wavelet transform to reduce
dimensionality before classification with the second-stage classifier.
In the \ac{cble}-decoder, it is this wavelet transform that decreases the sensitivity
to latency differences, actively compensating for \ac{erp} latency jitter.

@@

The W\ac{cble} algorithm is presented in Algorithm~\ref{alg:wcble}.
Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using equation~\ref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall \quad n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  Schematic representation of the \ac{wcble} training phase.
}{%
  Schematic representation of the \ac{wcble} training phase.
  (1) The first-stage spatiotemporal binary classifier is trained on a set of
  epochs.
  (2) It is	then applied to time-shifted copies of the epochs to obtain scores
  and class probabilities	over time.
  (3) The medians of these probability distributions are assumed the class
  latencies.
  (4) The epochs are aligned to their corresponding	class latencies by
  shifting in time such that all latencies fall at the same moment.
  (5) The spatiotemporal classifier is then retrained on the aligned epochs
  for a	next iteration.
  (6) After the iterative process halts, the scores and	latencies obtained
  from the last iteration are used to train the second-stage
  classifier.
}{%
  fig:diagram-train
}
The process halts after a fixed amount of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in Figure~\ref{fig:diagram-eval} to obtain a
vector of scores over time as in~Section~\ref{sec:cble} and the estimated latencies as
in~Section~\ref{sec:robust-latency}.
\fullpagefig{
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}

}{%
  Schematic representation of the \ac{wcble} test phase.
}{%
  Schematic representation of the Woody Classifier Based Latency
  Estimation test phase. (1) The
	first-stage	spatiotemporal binary classifier obtained from the training phase is
	applied to time-shifted copies of the epochs to obtain scores and class probabilities
	over time. (2) The medians of these probability distributions are assumed
	as the new class latencies. (3) The scores and class latencies are input to the
  trained second-stage classifier, which predicts the label of the epochs.
}{%
  fig:diagram-eval
}
\todo{re-insert comments}

\begin{algorithm}[H]
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
  \mathbf{l}^\mathrm{train}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\State $\theta \gets
			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l})$
      %\algorithmiccomment{Train stage 1}
		\For{$n=1\ldots N$}
      %\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\theta,
  \mat{Y}^\mathrm{train},
  \mathbf{s}_\mathrm{target}^\mathrm{train},
  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
	\smallskip \hrule \smallskip
	\textsc{Evaluate}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\For{$m=1\ldots M$}
      %\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\mat{Y}^\mathrm{test},
  \mathbf{s}_\mathrm{target}^\mathrm{test},
  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$

\begin{algorithm}[H]
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l})
		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
	\begin{algorithmic}[1]
		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
    N$
      %\algorithmiccomment{Train stage 1}
		\Repeat
		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l}$
		\For{$n=1\ldots N$}
    \State  $s_n \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
		\EndFor
    \Until{convergence or maximum iterations reached}
    \For{$n=1\ldots N$}%\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets
      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
    \EndFor

	\smallskip \hrule \smallskip
	\textsc{Evaluate}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\For{$m=1\ldots M$}%\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{m,s}^\mathrm{test} \gets
      f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\mat{Y}^\mathrm{test},
  \mathbf{s}_\mathrm{target}^\mathrm{test},
  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
  \caption{\ac{wcble}}
	\label{alg:wcble}
\end{algorithm}

\section{Results}
\label{sec:wcble/results}
\todo{refer to later chapter for results on real data}
\subsection{Data simulation}
To verify our approach, we simulated 16 channel \ac{eeg} data with standard
10-20 electrode positions using the \texttt{simulate\_evoked} method of the
MNE (version 1.8)~\cite{Gramfort2013} software package.
\Acp{erp} were generated by projecting a sine wave pulse source time course in a dipole in the
left hemisphere to the scalp electrodes using a boundary element method forward
model~\cite{Mosher1999} MNE's \texttt{fsaverage} source space and anatomy.
Pink temporal noise is generated by passing gaussian noise through an infinite
impulse response filter with transfer function
$\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
Pink spatial noise is also added, using a noise covariance matrix constructed from the
cosine distances between electrode positions and scaling it by the inverse of
the resulting spatial.
We refer to~\textcite{Gramfort2014} for implementation details.
%Evoked potentials are simulated for three different source time course waveforms,
%a gaussian, a morlet wavelet and a sinusoid pulse with variable latency,
%defined by the following formulas:
%\begin{equation}
%  f_\text{gaussian}(t,l) = a
%  \frac{1}{\varsigma\sqrt{2\pi}}e^{-\frac{1}{2}\frac{\left(t-l\right)^2}{\varsigma/2}}
%\end{equation}
%\begin{equation}
%  f_\text{morlet}(t,l) = ae^{-\left(\frac{t-l}{\frac{c}{2\pi f}}\right)^2}
%\end{equation}
The source time course is defined by the following function:
\begin{equation}
  s(t,l) =
  \begin{cases}
    a\sin\left(2\pi f\left(t-l\right)\right) & \text{if  $l-\frac{1}{2f} < t < l-\frac{1}{2f}$} \\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
over time $t$, with latency $l$ and amplitude $a=\num{1e{-7}}$V and frequency
$f=4$.
%The standard deviation that determines the width gaussian $\varsigma=0.1s$.
%For the morlet wavelet, the number of cycles $c=3$ and the frequency $f=4Hz$.
%For the sine pulse, the frequency $f=4$

@@

%A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-sine}.
%\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
%\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}
}{%
  Simulated target and non-target evoked data.
}{%
  Simulated target and non-target evoked data. Targets contain a sine wave
  pulse jittered by $\sigma$. 32 EEG channels with pink
  spatiotemporal noise were simulated using a forward head model, channel C3 is
  plotted here.
}{fig:wcble/sim-sine}
\todo{Properly align plots and label axes}
\todo{decrease tick size}

\subsection{Latency estimation}
\todo{refer to figure}
We compared our proposed \ac{wcble} algorithm to \ac{cble} on a latency
estimation task using the pearson correlation coefficient $\rho$\todo{equation} at different
jitter and \ac{snr} levels.
For both \ac{cble} and \ac{wcble}, target latencies where extracted in a
10-fold cross-validation scheme using the robust median latency method after
fitting the model.
\Ac{wcble}'s maximum iteration number was set to 64.
Results are presented in Figure~\ref{fig:wcble/results/latency}.
A 95\ confidence intervals on each measure was obtained through bootstrapping
with 1000 permutations.

\begin{figure}[H]
  \input{figures/wcble/fig_accuracy.pgf}
  \caption{caption}
  \label{fig:wcble/results/latency}
\end{figure}

\todo{properly label axes}
\todo{uniform fonts with rest of document}
\todo{correct width}
\todo{figure for data generation}

@@

Analogous to the previous experiment, 10-fold cross-validated classification
accuracy between target and non-target epochs was assessed.
Here, \ac{wcble} was compared to \ac{cble}, but both also to their base classifier
\ac{tlda}.
\Ac{cble} and \ac{wcble} classifiers were implemented as described
in~\ref{sec:wcble/methods}, with both the wavelet transformed score time series
and square transformed latencies as input to a second-stage logistic regression
classifier with $L_2$-norm regularization ($C=0.2$).
\todo{ref figure}

\begin{figure}
  \input{figures/wcble/fig_accuracy.pgf}
  \caption{caption}
  \label{fig:wcble/results/accuracy}
\end{figure}

The classification accuracy of both \ac{tlda} and \ac{cble} decreases with
$\$sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
This difference is particularily large when $\sigma$ is higher ($\sigma=0.2$
and $\sigma=0.3$) and \ac{snr} is not yet high.
\Ac{wcble} always scored better than \ac{cble} and \ac{wcble}, except for
$\sigma=0.1$ and $\text{\ac{snr}} <27\text{db}$, at which point both \ac{cble}
and \ac{wcble} performed at chance level.
When considering the accuracy of \ac{wcble}, we see that it shows more of a
plateau as \ac{snr} increases.
\todo{insert results figure and discuss}

@@

As shown, the proposed method is suitable for \ac{bci} decoding setting
as well as single-trial \ac{erp} latency analyses.
In the case where there is only one data class present, the first-stage
classifier can be replaced by a spatiotemporal filter.
Instead of outputting a classification score, this filter should now output a
metric representing the presence of a template response in the signal.
A suiting candidate is the spatiotemporal
\ac{lcmv}-beamformer~\cite{VanVliet2015}\todo{refer chapter}.
The activation pattern can be set to the average of the set of trials to
analyze, and it can be refined at each \ac{wcble} iteration by taking the
average of aligned trials.

@@

The main limitation of the proposed method is its supposed lack to handle
multi-component \ac{erp} data.
In real \ac{erp} analysis settings, multiple components are usually present,
and each component has a distinct contribution to class discriminatbility and
latency distribution.
A single-component method will yield issues in interpretability and convergence
in the presence of other components, as it can `lock-on' to a given component
cluster and flatten out the others.
This can be avoided, however, by the choice of a proper region of interest
isolating solely the component of interest.
Current work focuses on extending \ac{wcble} to a multi-component setting, on
the one hand to performance latency estimation and decoding performance on real
\ac{erp} data and on the other hand to apply it as an \ac{erp} component
separation method in the fashion of~\textcite{Oyang2017}.

@@

>>>>

\chapter{\Acs{erp} latency estimation and alignment}
\label{sec:wcble}
\emph{Sections~\ref{sec:wcble/literature/contrib} and~\ref{sec:wcble/methods} were published
as part of~\textcite{VanDenKerchove2024}}

@@

\Ac{cble}, summarized in \autoref{sec:wcble/app/cble-alg}, works by training a
\textit{first-stage} classifier $\mathcal{C}(\theta,f)$
defined within a time period $[s_1:s_2]$ with a set of parameters $\theta$ and a
decision function $f(\mat{X}[:,s_1:s_2],\theta) \rightarrow
y$ outputting a classification score $y\in\mathbb{R}$ for a given epoch $\mat{X}$,
such that
\begin{equation}
  \theta = \mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
  \label{eq:cble-train}
\end{equation}
Then,
%instead of applying the classifier directly to an unseen, cropped
%testing epoch $\mat{X}[:,s_1:s_2]$ and obtaining a single score value $y$,
$f$ can be applied to all (possibly overlapping) slices of length $s_2-s_1$ of
an epoch $\mat{X}$, resulting in a vector of score values
$\mathbf{y}=[y_1\,\ldots\,y_R]^T \in\mathbb{R}^R$ such that
\begin{equation}
  y_s = f(\mat{X}[:,s:s+(s_2-s_1)],\theta)\quad \forall s\in 1,\ldots,R
	\label{eq:score}
\end{equation}
with $R = S-(s_2-s_1)$.
To leverage \ac{cble} for \ac{erp} classification, the score vectors $\mathbf{y}$ can be
arranged in matrices $\mat{Y}^\mathrm{train}\in\mathbb{R}^{N\times R}$ and $\mat{Y}^\mathrm{test}\in\mathbb{R}^{M\times R}$.
These can be further classified by training a \textit{second-stage} classifier on
$\mat{Y}^\mathrm{train}$ and class labels $\mathbf{l}^\mathrm{train}$.
However, the resulting score-over-time vectors per epoch still suffer from jitter.
For classification, we follow the approach of \cite{Mowla2017}, using a
maximum-level hierarchical Daubechies-4 wavelet transform to reduce
dimensionality before classification with the second-stage classifier.
In the \ac{cble}-decoder, it is this wavelet transform that decreases the sensitivity
to latency differences, actively compensating for \ac{erp} latency jitter.

@@

The W\ac{cble} algorithm is presented in \autoref{sec:wcble/app/wcble-alg}.
Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using equation~\ref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall \quad n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  Schematic representation of the \ac{wcble} training phase.
}{%
  Schematic representation of the \ac{wcble} training phase.
  (1) The first-stage spatiotemporal binary classifier is trained on a set of
  epochs.
  (2) It is	then applied to time-shifted copies of the epochs to obtain scores
  and class probabilities	over time.
  (3) The medians of these probability distributions are assumed the class
  latencies.
  (4) The epochs are aligned to their corresponding	class latencies by
  shifting in time such that all latencies fall at the same moment.
  (5) The spatiotemporal classifier is then retrained on the aligned epochs
  for a	next iteration.
  (6) After the iterative process halts, the scores and	latencies obtained
  from the last iteration are used to train the second-stage
  classifier.
}{%
  fig:diagram-train
}
The process halts after a fixed amount of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in Figure~\ref{fig:diagram-eval} to obtain a
vector of scores over time as in~Section~\ref{sec:cble} and the estimated latencies as
in~Section~\ref{sec:robust-latency}.
\fullpagefig{
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}

}{%
  Schematic representation of the \ac{wcble} test phase.
}{%
  Schematic representation of the Woody Classifier Based Latency
  Estimation test phase. (1) The
	first-stage	spatiotemporal binary classifier obtained from the training phase is
	applied to time-shifted copies of the epochs to obtain scores and class probabilities
	over time. (2) The medians of these probability distributions are assumed
	as the new class latencies. (3) The scores and class latencies are input to the
  trained second-stage classifier, which predicts the label of the epochs.
}{%
  fig:diagram-eval
}
\section{Results}
\label{sec:wcble/results}
Since it is hard to obtain ground truth measures of \ac{erp} latencies, we
first evaluate our approach with synthetic data.
In \autoref{sec:covert-align}, the proposed algorithm will be applied to real
\ac{eeg} data.

\subsection{Synthetic data generation}
To verify our approach, we simulated synthetic 16 channel \ac{eeg} data with standard
10-20 electrode positions using the \texttt{simulate\_evoked} method of the
MNE (version 1.8)~\cite{Gramfort2013} software package.
\Acp{erp} were generated by projecting a sine wave pulse source time course in a dipole in the
left hemisphere to the scalp electrodes using a boundary element method forward
model~\cite{Mosher1999} MNE's \texttt{fsaverage} source space and anatomy.
Pink temporal noise is generated by passing gaussian noise through an infinite
impulse response filter with transfer function
$\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
Pink spatial noise is also added, using a noise covariance matrix constructed from the
cosine distances between electrode positions and scaling it by the inverse of
the resulting spatial.
We refer to~\textcite{Gramfort2014} for implementation details.
%Evoked potentials are simulated for three different source time course waveforms,
%a gaussian, a morlet wavelet and a sinusoid pulse with variable latency,
%defined by the following formulas:
%\begin{equation}
%  f_\text{gaussian}(t,l) = a
%  \frac{1}{\varsigma\sqrt{2\pi}}e^{-\frac{1}{2}\frac{\left(t-l\right)^2}{\varsigma/2}}
%\end{equation}
%\begin{equation}
%  f_\text{morlet}(t,l) = ae^{-\left(\frac{t-l}{\frac{c}{2\pi f}}\right)^2}
%\end{equation}
The source time course is defined by the following function:
\begin{equation}
  s(t,l) =
  \begin{cases}
    a\sin\left(2\pi f\left(t-l\right)\right) & \text{if  $l-\frac{1}{2f} < t < l-\frac{1}{2f}$} \\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
over time $t$, with latency $l$ and amplitude $a=\num{1e{-7}}$V and frequency
$f=4$.
%The standard deviation that determines the width gaussian $\varsigma=0.1s$.
%For the morlet wavelet, the number of cycles $c=3$ and the frequency $f=4Hz$.
%For the sine pulse, the frequency $f=4$

@@

%A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-sine}.
%\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
%\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
\fullpagefig{%
  %\includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}
  \sffamily
  \sansmath
  \inputpgf{figures/wcble}{simulated-sine.pgf}
}{%
  Simulated target and non-target evoked data.
}{%
  Simulated target (bottom half of epochs) and non-target (top half of epoch) evoked data
  and their average (red: target, yellow: noon-target, shaded are: standard
  deviation).
  Targets contain a sine wave pulse jittered by $\sigma$. 32 EEG channels with pink
  spatiotemporal noise were simulated using a forward head model, channel C3 is
  plotted here.

}{fig:wcble/sim-sine}
\todo{label x axis}

\subsection{Latency estimation}
We compared our proposed \ac{wcble} algorithm to \ac{cble} on a latency
estimation task using the pearson correlation coefficient $\rho$ at different
jitter and \ac{snr} levels.
For both \ac{cble} and \ac{wcble}, target latencies where extracted in a
10-fold cross-validation scheme using the robust median latency method after
fitting the model.
\Ac{wcble}'s maximum iteration number was set to 64.
Results are presented in Figure~\ref{fig:wcble/results/latency}.
A 95\ confidence intervals on each measure was obtained through bootstrapping
with 1000 permutations.

\begin{figure}
  \sffamily\sansmath
  \input{figures/wcble/fig_latency.pgf}
  \caption[Synthetic latency estimation results]{Correctness of estimated
  latency for varying jitter ($\sigma$) and \ac{snr} introduced in synthetic evoked target
  target data, measured as pearson correlation coefficient $\rho$ between
  estimated and ground truth latencies.
  \Ac{wcble} estimates latencies more accurately in the presence of noise and
  high jitter.}
  \label{fig:wcble/results/latency}
\end{figure}

@@

Analogous to the previous experiment, 10-fold cross-validated classification
accuracy between target and non-target epochs was assessed.
Here, \ac{wcble} was compared to \ac{cble}, but both also to their base classifier
\ac{tlda}.
\Ac{cble} and \ac{wcble} classifiers were implemented as described
in~\ref{sec:wcble/methods}, with both the wavelet transformed score time series
and square transformed latencies as input to a second-stage logistic regression
classifier with $L_2$-norm regularization ($C=0.2$).

\begin{figure}
  \sffamily\sansmath
  \input{figures/wcble/fig_accuracy.pgf}
  \caption[Synthetic decoding results]{Decoding performance, measured as binary
  classification accuracy, for \ac{cble}, \ac{wcble} and their first stage
  classifier \ac{tlda} for varying introduced jitter ($\sigma$) and \ac{snr} in
  synthetic evoked target and non-target data.
  \Ac{wcble} decoding performance is robust to more noise and jitter than the other methods.}
  \label{fig:wcble/results/accuracy}
\end{figure}

\autoref{fig:wcble/results/accuracy} shows that the classification accuracy of both \ac{tlda} and \ac{cble} decreases with
$\$sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
This difference is particularily large when $\sigma$ is higher ($\sigma=0.2$
and $\sigma=0.3$) and \ac{snr} is not yet high.
\Ac{wcble} always scored better than \ac{cble} and \ac{wcble}, except for
$\sigma=0.1$ and $\text{\ac{snr}} <27\text{db}$, at which point both \ac{cble}
and \ac{wcble} performed at chance level.
When considering the accuracy of \ac{wcble}, we see that it shows more of a
plateau as \ac{snr} increases.

@@

As shown, the proposed method is suitable for \ac{bci} decoding setting
as well as single-trial \ac{erp} latency analyses.
In the case where there is only one data class present, the first-stage
classifier can be replaced by a spatiotemporal filter.
Instead of outputting a classification score, this filter should now output a
metric representing the presence of a template response in the signal.
A suiting candidate is the spatiotemporal
\ac{lcmv}-beamformer~\cite{VanVliet2015} described in
\autoref{sec:stbf-struct}.
The activation pattern can be set to the average of the set of trials to
analyze, and it can be refined at each \ac{wcble} iteration by taking the
average of aligned trials.

@@

The main limitation of the proposed method is its supposed lack to handle
multi-component \ac{erp} data.
In real \ac{erp} analysis settings, multiple components are usually present,
and each component has a distinct contribution to class discriminatbility and
latency distribution.
A single-component method will yield issues in interpretability and convergence
in the presence of other components, as it can `lock-on' to a given component
cluster and flatten out the others.
This can be avoided, however, by the choice of a proper region of interest
isolating solely the component of interest.
Current work focuses on extending \ac{wcble} to a multi-component setting, on
the one hand to performance latency estimation and decoding performance on real
\ac{erp} data and on the other hand to apply it as an \ac{erp} component
separation method in the fashion of~\textcite{Ouyang2017}.

@@

\clearpage
\begin{subappendices}
\section{\Acl{cble} algorithm}
\label{sec:wcble/app/cble-alg}
\begin{algorithm}[h]
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
  \mathbf{l}^\mathrm{train}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\State $\theta \gets
			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l})$
      \algorithmiccomment{Train stage 1}
		\For{$n=1\ldots N$}
      \algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\theta,
  \mat{Y}^\mathrm{train},
  \mathbf{s}_\mathrm{target}^\mathrm{train},
  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
	\smallskip \hrule \smallskip
	\textsc{Evaluate}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\For{$m=1\ldots M$}
      \algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\mat{Y}^\mathrm{test},
  \mathbf{s}_\mathrm{target}^\mathrm{test},
  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
